---
title: "ED models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ED models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: ref.bib
link-citations: true
link-bibliography: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette provides instructions on how to construct rate matrices ($Q$s) for Embedded Dependency (ED) Models suitable for phylogenetic inference with anatomically dependent (inapplicable) morphological characters [@tarasov2022new]. You can copy-paste the constructed matrices into [RevBayes](https://revbayes.github.io) scripts or integrate them with other R packages.

# Install rphenoscate package

[rphenoscaTe](https://github.com/uyedaj/rphenoscate) provides functions for creating ED models. [rphenoscapeTe](https://github.com/uyedaj/rphenoscate) depends on [rphenoscaPe](https://rphenoscape.phenoscape.org). Both packages are under active development and will soon be submitted to CRAN. Currently, they are available via github. Make sure that you use the most recent version of R (>=4.1).

```{r eval=F, echo=T}
# Requires R>=4.1 !!!
install.packages("remotes")
library(remotes)
install_github("phenoscape/rphenoscape")
install.packages("numbers")
install_github("uyedaj/rphenoscate")
library(rphenoscate)

```


# Constructing Character Hierarchies: the DDA algorithm

To construct a rate matrix for a given character hierarchy, it is necessary first to draw, for example on paper, a dependency diagram (for example, Figure 1B-E) reflecting the hierarchical relationship between characters, and then manually apply the DDA (Dependency Diagram Amalgamation) algorithm to create the amalgamated matrix.

A dependency diagram consists of three building blocks (Fig. 1F): (i) absent/present (*a/p*) character; (ii) qualitative character; and (iii) dependency arrow. In the case of the *a/p* character, there are two subblocks: the *a/p-node* and the *p-node*, which are the character itself and its state *p*, respectively. In the diagram, the *a/p* and qualitative characters are interlinked by means of a dependency arrow to reflect hierarchical relationships. Specifically, the dependency arrow links the *p-node* of a controlling character with a dependent character (which can be either of the *a/p* or quantitative type).

The DDA algorithm (Fig. 1B) works by taking the diagram and traversing it from tips to root in a topologically ordered manner. As you traverse, you need to amalgamate characters at *a/p-* or *p-nodes* according to the following rules:

1.  At each *p-node* combine all child characters via SMM amalgamation [the equations (2) and (3) in @tarasSMM].
2.  At each *a/p-node* conditionally combine all child characters via:
    -   *ED-ql* amalgamation if the children are qualitative characters.
    -   *ED-bd* amalgamation if the children are *a/p* characters.
    -   The mixture of *ED-ql* and *ED-bd* amalgamations if the children are both *a/p* and qualitative characters. This requires appropriate construction of the embedded dependency vector ($\boldsymbol{\phi_{d}}$) to match the dependencies between states.
3.  The amalgamation at the root of the dependency diagram yields the desired final rate matrix.

The R code below is provided for manual amalgamations with the DDA algorithm for the four examples shown in Figure 1.

![Fig. 1. **A)** Complex hierarchy of three *a/p* and four qualitative characters. **B)** Dependency Diagram for (A); "\#" indicates the sequence of steps in the DDA algorithm; *SMM()* idicates SMM amalgamation; *ED-bd()* and *ED-ql()* indicate *a/p* and qualitative types of the ED amalgamations respectively. **C-D)** Dependency Diagrams for the tail characters from @tarasov2022new.](https://github.com/sergeitarasov/MorphoModels/blob/main/vignettes/Fig_1_vignette.png){width="90%"}


## Example 1. Complex Hierrachy from Fig. 1B

This amalgamation assumes that qualitative and a/p characters evolve at different rates (indicated by symbols 2 and 1, respectively). We amalgamate three a/p and four qualitative characters, and print the $Q$ matrix to be used in [RevBayes](https://revbayes.github.io).

```{r message=FALSE, warning=FALSE}
library(rphenoscate)
```

```{r echo=T, results='hide'}
# STEP #1
# Q can be defined as
C1 <-matrix(c(-1, 1, 1, -1), 2, 2, byrow = TRUE, dimnames =list( c("r", "b"), c("r", "b")) )
# or using the special function initQ
C1 <- initQ(c('r', 'b'), 1)
S <- initQ(c('o', 'c'), 1)
step1 <- amaSMM( C1, S)

# STEP #2
# armor: A* absent, A present
A <- initQ(c('A*', 'A'), 2)
step2 <- amaED(A, step1, type = "ql")

# STEP #3
C2 <- initQ(c("g", "p"), 1)

# STEP #4
# spine: S* absent, S present
Sp <- initQ(c('S*', 'S'), 2)
step4 <- amaED(Sp, C2, type = "ql")

# STEP #5
# size
Si <- initQ(c("l", "s"), 1)
# !!! For automatic construction of phi in step 6, the order of SMM amalgamation is important: first place a/p characters in amaSMM() and only then the qualitative ones.
step5 <- amaSMM(step2, step4, Si)

# STEP #6
# tail: T* absent, T present
Tl <- initQ(c('T*', 'T'), 2)
step6 <- amaED(Tl, step5, type = "ap")

# OR with manual phi
phi <- c(1,1, rep(0, 28))
step6 <- amaED(Tl, step5, phi=phi)

dim(step6)

# to RevBayes
Qrb1 <- as_matrixRB(step6)
Qrb2 <- as_matrixRB(step6, symb = 'q')

# copy/paste this matrices into RevBayes script
cat(Qrb1)
cat(Qrb2)

# or save
#cat(Qrb2, file='Qrb2.txt')
```

## Example 2. Tail Color Case (Fig. 1C)

```{r}
C <- initQ(c("r", "b"), 1)
Tl <- initQ(c('T*', 'T'), 1)

Qtc <- amaED(Tl, C, type = "ql")
cat(as_matrixRB(Qtc))
```

## Example 3. Tail Armor Case (Fig. 1D)

```{r}
C1 <- initQ(c("r", "b"), 1)
A <- initQ(c('A*', 'A'), 2)
Tl <- initQ(c('T*', 'T'), 2)

step1 <- amaED(A, C, type = "ql")
Qta <- amaED(Tl, step1, type = "ap")
cat(as_matrixRB(Qta, symb='q'))
```

## Example 4. Tail Color + Tail Color Case (Fig. 1E)

```{r}
C1 <- initQ(c("r", "b"), 1)
A <- initQ(c('A*', 'A'), 2)
C <- initQ(c("b", "r"), 1)
Tl <- initQ(c('T*', 'T'), 2)

step1 <- amaED(A, C, type = "ql")
step2 <- amaSMM(step1, C)
Qac <- amaED(Tl, step2 , type = "ap")
cat(as_matrixRB(Qac, symb='q'))
```

# References
